---
title: "HW"
author: "Lynn Waterhouse"
date: "March 29, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment

This homework includes problems that may require: T-tests, chi-square, linear regression, or multiple linear regression.

For each problem:

(1) Read in the data
(2) choose the correct test.
(3) write out the null and alternative hypotheses
(4) verify the assumptions are met
(5) conduct the test and report p-value
(6) Write out the result of the test.
(7) Write out a conclusion from the result.

For some problems you will also be asked to make a plot.


## Problem 1

We conduct a survey and count the number of individuals for 6 different fish species from inside a marine reserve and from outside the marine reserve. We want to know if there is a difference.  The data are found in "marinepark.csv".

Make a 2 panel barplot (use par(mfrow=c(2,1))) and plot the # of species for Inside the Marine Park and for Outside the Marine Park.

```{r echo=FALSE}
set.seed(123)
outside<-c(54,79,30,81,1005,720)
inside<-round(outside*runif(6,.8,2.3))
mydat<-data.frame(rbind(inside,outside))
colnames(mydat)<-c("Grouper","Ray","Shark","Snapper","Wrasse","Jack")
write.csv(mydat,"marinepark.csv")
```

```{r eval=FALSE, echo=FALSE}
#(1) Read in the data
mydat<-read.csv("marinepark.csv")
#(2) write out the null and alternative hypotheses
#    H0: the location and species composition are independent
#    Ha: the location and species composition are not independent
#(4) choose the correct test. 
#    chi-square test for independence
#(3) verify the assumptions are met
#   all categories >5
#(4) conduct the test and report p-value
chisq.test(mydat)
#   p-value < 2.2e-16
#(5) Write out the result of the test.
#   reject the null hypothesis
#(6) Write out a conclusion from the result.
#   we conclude that location and species composition are not indepedent, 
#   there is some reason other than random chance that there are a different # of 
#   each species in the marine park and outside the marine park
par(mfrow=c(2,1),oma=c(.1,.1,.1,.1), mar=c(2,3,1,1))
barplot(inside,ylab="Count",main="Inside Marine Park")
barplot(outside,names.arg=colnames(mydat),ylab="Count",main="Outside Marine Park")

```

##Problem 2
Every morning for two months we swim a transect and record the number of eels that we see. We do this in two locations. Inside and outside a marine park. We want to know if we see more eels in the marine park than outside.  One of the assumptions will be that both sets of counts are normally distributed. The data are in "eels.csv".  

Make 1 boxplot showing the data side by side for inside and outside the marine park.

```{r,echo=FALSE}
set.seed(122)
location<-c(rep("inside",60),rep("outside",60))
count<-c(round(rnorm(60,5,2)),rbinom(60,10,.4))
eel.dat<-data.frame(location=location,count=count)
write.csv(eel.dat,"eels.csv")
```

```{r,eval=F,echo=F}
#(1) Read in the data
eel<-read.csv("eels.csv")
inside<-eel.dat$count[which(eel.dat$location == "inside")]
outside<-eel.dat$count[which(eel.dat$location == "oustide")]
#(2) write out the null and alternative hypotheses
#    H0: mean # eels inside < = mean # eels outside
#    Ha: mean # eels inside > mean # eels oustide
#(4) choose the correct test. 
#   2 sample T-test, one sided
var.test(inside,outside) #p-value = 0.0001007, nonhomogeneous variances
#(3) verify the assumptions are met
hist(inside)
hist(outside)
shapiro.test(inside)
shapiro.test(outside)
qqnorm(inside)
qqline(inside)
qqnorm(oustide)
qqline(outside)
#  we assume independent samples and both samples are normally distributed
#(4) conduct the test and report p-value
t.test(inside,outside,alternative="greater",var.equal=FALSE)
#  p-value = 0.1727
#(5) Write out the result of the test.
#  p-value = 0.1727>0.05, fail to reject NULL
#(6) Write out a conclusion from the result.
#    There is not enough information to conclude that there are statistically
#   significantly more eels inside the marine reserve than outside.

#boxplot
boxplot(count~location,data=eel.dat,ylab="Count Eels (#)",main="Number Eels Inside \n& Outside Marine Park")
```

##Problem 3
We are curious as to if sea surface temperature or abundance of prey fish or the interaction between the two or some combination of the two are a better predictor of tuna abundance. Read in the data "tunacount.csv".

Plot the raw data and the best fit result.
```{r,echo=FALSE}
set.seed(2)
temp<-runif(30,17.1,18)
sst<-round(c(temp,temp*runif(30,1.1,1.2),temp*runif(30,1.1,1.3)),2)
prey<-rbinom(90,5000,.8)
tuna<-NULL
for (i in 1:90){
  tuna[i]<-round(rnorm(1,mean=-3+.2*sst[i]+prey[i]*.0005+.001*sst[i]*prey[i],sd=1))
}
mydat<-data.frame(sst=sst,prey=prey,tuna=tuna)
write.csv(mydat,"tunacount.csv")
```
