---
title: "Poisson GLM"
date: "March 29, 2017"
output: html_document
---

```{r eval=FALSE, echo=FALSE}
# Generate data
# survey data from 3 sites
set.seed(100)
sites <- c("costa", "bahia", "marietas")
n.sites <- length(sites)
N.vec <- round(rnorm(n.sites, mean=200, sd=10))
true.dens <- c(6, 10, 6.5)
mantas <- vector("list",n.sites)
for(i in 1:n.sites) mantas[[i]] <- rpois(N.vec[i], lambda=true.dens[i])
# lapply(mantas,mean)

dat <- data.frame(site = rep(sites, N.vec),
                  counts = unlist(mantas))
write.csv(dat,"/home/brian/Documents/Workshops/Mexico fisheries/StatsClass/poissonGLM.csv",row.names=FALSE)
```

## Review: Poisson distribution
Common distribution to model **count data**: the number of individuals, arrivals, events, etc., in a given time/space/unit of effort if each event is independent of all the others.

#### Ex1. You survey a location for the number of mantas you see in 2 hours. You think each manta sighting is independent. If the true mean density (# of mantas seen per 2 hours) is 8, plot the probability density function of seeing between 0 and 20 mantas.
```{r}
x <- 0:20
dvals <- dpois(x,lambda=8)
plot(x,dvals,type='l')
```

You see 5 mantas in your 2-hour survey, less than the mean (= 8). Unlucky!

```{r echo=FALSE}
plot(x,dvals,type='l')
abline(v=5,col='red',lwd=4)
```

What is the probability any random 2-hour survey would see **more** than you did?

```{r}
ppois(5,lambda=8,lower.tail=FALSE)
```

## Poisson GLM 
If your response variable is count data, then a Poisson GLM is probably more appropriate than ANOVA or simple linear regression **because they assume your data is normally distributed**. In Ex1, we could use `dpois` and `ppois` because we knew the true density. We will never know the true density, and we will want to *estimate it using our count data*.

#### Poisson GLM estimates density from count data (surveys)

#### Ex2. Manta surveys by site.
You conduct about 150 manta surveys in 3 locations. Each data point is the number of mantas you see in 2 hours. You want to know if the density of mantas is different by site.

Looks like ANOVA (3 sites). Let's plot our data and then test ANOVA assumptions.
```{r results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(multcomp)
# read in data
setwd("/home/brian/Documents/Workshops/Mexico fisheries/StatsClass")
dat <- read.csv("poissonGLM.csv", header=TRUE)
# look at data
head(dat)
names(dat)
dim(dat)
summary(dat)
```

```{r}
# boxplot of counts by site
boxplot(counts ~ site, data=dat,
        xlab="Site",
        ylab="Counts")
```

```{r}
# plot each site's distribution of counts
bahia <- filter(dat, site=="bahia")$counts
costa <- filter(dat, site=="costa")$counts
marietas <- filter(dat, site=="marietas")$counts

hist(bahia)
hist(costa)
hist(marietas)
```

ANOVA assumes each site is normally distributed. Look at QQ plots:
```{r}
par(mfrow = c(2,2), mar=c(2, 4, 1, 1) + 0.1)
qqnorm(bahia, main="bahia")
qqline(bahia)
qqnorm(costa, main="costa")
qqline(costa)
qqnorm(marietas, main="marietas")
qqline(marietas)
```

2 problems with histograms / QQ plots:

- discrete data ("steps")
- "bahia" looks ok but "marietas" and "costa" skewed right

#### Fit a Poisson GLM
```{r}
pois.mod <- glm(counts ~ site, data=dat, family=poisson)
summary(pois.mod)
```

#### How do we interpret this?
The GLM takes `bahia` as the mean, and has estimated deviations from `bahia` as the effects of `costa` and `marietas`.

`sitecosta` has `Estimate` = `-0.50418` and `Pr(>|z|)` `<2e-16 ***`: The density of mantas on the costa is **significantly lower** than in bahia.

`sitemarietas` has `Estimate` = `-0.44121` and `Pr(>|z|)` `<2e-16 ***`: The density of mantas at the marietas is **significantly lower** than in bahia.

#### What if we want to test if there is a difference between Costa and Marietas?
We can do a Tukey test on the GLM output just like we did with ANOVA output. Use the same `glht` function from the `multcomp` package:
```{r}
pois.tukey <- glht(pois.mod, mcp(site="Tukey"))
summary(pois.tukey)
```

So our data do NOT show a difference between Costa and Marietas, but we do show Bahia has higher density of mantas than both Costa and Marietas.

#### Calculate mean density for each site
Do this by using the `predict` function with `type="response"`:
```{r}
newdata <- data.frame(site=c("bahia","costa","marietas"))
predict(pois.mod, newdata, type="response")
```

#### Get 95% CI for density estimates by site
```{r}
pred <- predict(pois.mod, newdata, type="response", se.fit=TRUE)
lwr <- pred$fit - 1.96*pred$se.fit
upr <- pred$fit + 1.96*pred$se.fit
```

So our estimates of mean density and 95% CI are:

- Bahia: 10.1 (9.6 -- 10.5)
- Costa: 6.1 (5.7 -- 6.4)
- Marietas: 6.5 (6.1 -- 6.8)

#### The Poisson GLM fits better than ANOVA/normal (lower AIC)
Also see that the Normal GLM is identical to an ANOVA in this case.
```{r}
aov.mod <- aov(counts ~ site, data=dat)
norm.mod <- glm(counts ~ site, data=dat, family=gaussian)
AIC(pois.mod, aov.mod, norm.mod)
```

